Reduced plasticity and variance in physiological rates of ectotherm populations under climate change

### 2024-02-16 ms.qmd
Fix some spelling errors and formatting Supp

#### Affliations:
^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600, Australia &NewLine;  
^2^ Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW, Australia &NewLine;  
^3^ School of Life and Environmental Sciences, University of Sydney, Sydney, NSW 2006, Australia  &NewLine;  
^4^ Department of Biology, Lancaster University, Liverpool, UK &NewLine;  
$\dagger$ contributed equally &NewLine;  
$\ddagger$ corresponding author, daniel.noble@anu.edu.au &NewLine;

@@

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expected to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. Such responses depend on genetic and phenotypic variability. However, most work has focused on changes in mean phenotypic responses to climate warming ignoring how temperature may also change phenotypic variability. Phenotypic variability may be particularly important at extreme, high temperatures, which would facilitate selection of resistant individuals or promote plasticity (acclimation) and thereby increase resilience to heat waves. Using newly developed effect size estimates and meta-analysis (>1900 effects from 226 species), we show that across habitats relative variance in physiological rates decreased at higher temperatures. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures, but capacity for plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Neither life-history stage nor past climate explained effect variability. Our results show that beneficial acclimation responses may trade-off with reductions in physiological rate variance. This trade-off could constrain evolutionary responses to climate change and reduce the potential benefits of portfolio effects. These findings have important evolutionary and ecological ramifications that affect our understanding of how climate change will impacts populations now and in the future.

@@

Periods of past climatic change have had disproportionate impacts on some ecosystems over others raising the question of which ecosystems will be most vulnerable to contemporary climate change. Species occupying terrestrial ecosystems are are thought to be particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper]. However, this conclusion has been questioned given that marine ectotherms have recently been shown to be closer to their upper thermal limits compared to terrestrial species [@Pinsky2019-fn]. Marine and freshwater ecosystems appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, it is unclear if the magnitude of physiological adjustment is sufficient to compensate for potentially negative environmental effects, particularly when temperature interacts with other abiotic changes. Low oxygen availability may be a major factor influencing the vulnerability of aquatic ecosystems, whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms which may have important cascading effects on energy transfer and productivity across different ecosystems [@seebacher2023can; @barneche2021warming].

@@

Using a large database on physiolgical rates[@Seebacher2015-wi] for marine, freshwater and terrestrial ectotherms we apply new effect size estimates [@Noble2022-ty] that capture changes in mean physiological rates ($lnRR_{Q_{10}}$), and changes in their relative variance ($lnCVR_{Q_{10}}$), standardised to a 10&deg;C temperature difference. These standardised effects sizes can be converted to percentage differences to permit comparisons between the relative magnitudes of rate changes resulting from acclimation and relative variance changes at high temperatures. Acclimation responses can be determined by comparing acclimation $lnRR_{Q_{10}}$ ($lnRR_{Q_{10_{acclimation}}}$) to acute $lnRR_{Q_{10}}$ ($lnRR_{Q_{10_{acute}}}$) from ectotherms acclimated to different temperatures. The difference in $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ measures the extent of acclimation (i.e., plasticity).

@@

## **Materials and Methods**
### *Literature collection*
We compiled literature on ectothermic animals that measured physiological rates (e.g., metabolic rate) at two or more temperatures after having been acclimated (or acclimatized) at these temperatures. We used data from a previous meta-analysis [@Seebacher2015-wi] and updated @Seebacher2015-wi's data by extracting data from suitable studies from our own searches that followed the same search protocol. More specifically, we performed a literature search using the Web of Science database for articles or proceedings papers published in English from 2013 to 2017 (the date after [@Seebacher2015-wi] searches were conducted) using the following topic search string: *"(acclimat* AND (therm* OR temp*) NOT (plant* OR tree* OR forest* OR fung* OR mammal* OR marsup* OR bird* OR human OR exercis* OR train* OR hypoxi*))"*. We further limited to the following research areas: Anatomy Morphology; Biodiversity Conservation; Biology; Ecology; Endocrinology Metabolism; Entomology; Evolutionary Biology; Marine Freshwater Biology; Physiology; Respiratory System, Reproductive Biology, Zoology.

@@

Lastly, to illustrate the effects that climate warming could have on physiological rate variance we also extracted climate projections and calculated thermal variability and predictability for the future. We used the CanESM2 climate model (2005-2100) under a high emissions scenerio (RCP8.5).

@@

@@

```{r, tbl-s1}
#| label: tbl-s1
#| echo: false
#| tbl-cap: Model estimates, standard error, and 95% credible intervals comparing changes in acute and acclimation $lnVR_{Q_{10}}$ across habitat types. Models estimates are based off 1,253 effect sizes from 139 studies.

@@

#### Publicaton Bias Analysis

Funnel plots did not show any noticable deviation from the typical funnel shape for any of the effect size estimates (@fig-s2).

>>>>

#### Affiliations:
^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600, Australia &NewLine;  
^2^ Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW, Australia &NewLine;  
^3^ School of Life and Environmental Sciences, University of Sydney, Sydney, NSW 2006, Australia  &NewLine;  
^4^ Department of Biology, Lancaster University, Liverpool, UK &NewLine;  
$\dagger$ contributed equally &NewLine;  
$\ddagger$ corresponding author, daniel.noble@anu.edu.au &NewLine;

@@

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expected to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. Such responses depend on genetic and phenotypic variability. However, most work has focused on changes in mean phenotypic responses to climate warming ignoring how temperature may also change phenotypic variability. Phenotypic variability may be particularly important at extreme, high temperatures, which would facilitate selection of resistant individuals or promote plasticity (acclimation) and thereby increase resilience to heat waves. Using newly developed effect size estimates and meta-analysis (>1900 effects from 226 species), we show that across habitats relative variance in physiological rates decreased at higher temperatures. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures, but capacity for plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Neither life-history stage nor past climate explained effect variability. Our results show that beneficial acclimation responses may trade-off with reductions in physiological rate variance. This trade-off could constrain evolutionary responses to climate change and reduce the potential benefits of portfolio effects. These findings have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

Periods of past climatic change have had disproportionate impacts on some ecosystems over others raising the question of which ecosystems will be most vulnerable to contemporary climate change. Species occupying terrestrial ecosystems are thought to be particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper]. However, this conclusion has been questioned given that marine ectotherms have recently been shown to be closer to their upper thermal limits compared to terrestrial species [@Pinsky2019-fn]. Marine and freshwater ecosystems appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, it is unclear if the magnitude of physiological adjustment is sufficient to compensate for potentially negative environmental effects, particularly when temperature interacts with other abiotic changes. Low oxygen availability may be a major factor influencing the vulnerability of aquatic ecosystems, whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms which may have important cascading effects on energy transfer and productivity across different ecosystems [@seebacher2023can; @barneche2021warming].

@@

Using a large database on physiological rates[@Seebacher2015-wi] for marine, freshwater and terrestrial ectotherms we apply new effect size estimates [@Noble2022-ty] that capture changes in mean physiological rates ($lnRR_{Q_{10}}$), and changes in their relative variance ($lnCVR_{Q_{10}}$), standardised to a 10&deg;C temperature difference. These standardised effects sizes can be converted to percentage differences to permit comparisons between the relative magnitudes of rate changes resulting from acclimation and relative variance changes at high temperatures. Acclimation responses can be determined by comparing acclimation $lnRR_{Q_{10}}$ ($lnRR_{Q_{10_{acclimation}}}$) to acute $lnRR_{Q_{10}}$ ($lnRR_{Q_{10_{acute}}}$) from ectotherms acclimated to different temperatures. The difference in $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ measures the extent of acclimation (i.e., plasticity).

@@

## **Methods & Protocols**
### *Literature collection*
We compiled literature on ectothermic animals that measured physiological rates (e.g., metabolic rate) at two or more temperatures after having been acclimated (or acclimatized) at these temperatures. We used data from a previous meta-analysis [@Seebacher2015-wi] and updated @Seebacher2015-wi's data by extracting data from suitable studies from our own searches that followed the same search protocol. More specifically, we performed a literature search using the Web of Science database for articles or proceedings papers published in English from 2013 to 2017 (the date after [@Seebacher2015-wi] searches were conducted) using the following topic search string: *"(acclimat* AND (therm* OR temp*) NOT (plant* OR tree* OR forest* OR fung* OR mammal* OR marsup* OR bird* OR human OR exercis* OR train* OR hypoxi*))"*. We further limited to the following research areas: Anatomy Morphology; Biodiversity Conservation; Biology; Ecology; Endocrinology Metabolism; Entomology; Evolutionary Biology; Marine Freshwater Biology; Physiology; Respiratory System, Reproductive Biology, Zoology.

@@

Lastly, to illustrate the effects that climate warming could have on physiological rate variance we also extracted climate projections and calculated thermal variability and predictability for the future. We used the CanESM2 climate model (2005-2100) under a high emissions scenario (RCP8.5).

@@

## Data Availability 
All data and code used to reproduce all analyses can be found on GitHub at: https://github.com/daniel1noble/Q10_meta_analysis.

@@

```{r, tbl-s1}
#| label: tbl-s1
#| echo: false
#| tbl-cap: Model estimates, standard error, and 95% credible intervals comparing changes in acute and acclimation $lnVR_{Q_{10}}$ across habitat types. Model estimates are based off 1,253 effect sizes from 139 studies.

@@

#### Publication Bias Analysis

Funnel plots did not show any noticeable deviation from the typical funnel shape for any of the effect size estimates (@fig-s2).
